---
title: "NYPD Shooting Incidents"
output: beamer_presentation
theme: "Madrid"
colortheme: "seahorse"
date: "2024-07-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r add_libraries, echo=FALSE,include=FALSE}
library(tidyverse)
library(lubridate)
```

```{r read_csv, echo=FALSE, include=FALSE}
shooting_data <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
#shooting_data
```

```{r remove_dups, echo=FALSE, include=FALSE}
shooting_data <- shooting_data %>% distinct(shooting_data$INCIDENT_KEY, shooting_data$OCCUR_TIME, .keep_all = TRUE)
#shooting_data
```

```{r clean, echo=FALSE, include=FALSE}
shooting_location <- shooting_data %>%
  select(c(OCCUR_DATE, BORO)) %>%
  mutate(OCCUR_DATE = as.Date(OCCUR_DATE, format = "%m/%d/%Y"), incidents = 1) %>%
  rename(Borough = "BORO") %>%
  drop_na()
#shooting_location
```

```{r summarize, echo=FALSE, include=FALSE}
#summary(shooting_location)
```

```{r group, echo=FALSE, include=FALSE}
shootings_per_month <- shooting_location %>% 
  group_by(month = floor_date(OCCUR_DATE, 'month'),Borough) %>%
  summarize(incidents=sum(incidents)) %>%
  select(month, Borough, incidents) %>%
  ungroup()
shootings_per_month
```

```{r by_borough, echo=FALSE, include=FALSE}
shootings_by_borough <- shooting_location %>%
  group_by(Borough) %>%
  summarize(incidents = sum(incidents)) %>%
  select(Borough, incidents) %>%
  ungroup()
```

```{r pop_csv, echo=FALSE, include=FALSE}
nyc_pop <- read_csv("https://data.cityofnewyork.us/api/views/xywu-7bv9/rows.csv?accessType=DOWNLOAD")
pop_2010 <- nyc_pop %>%
  select(c(Borough, '2010')) %>%
  filter(Borough != "NYC Total") %>%
  mutate(across(where(is.character), toupper)) %>%
  rename(population = '2010')
pop_2010
```

```{r shootings_2010, echo=FALSE, include=FALSE}
shootings_2010 <- shooting_location %>%
  group_by(year = year(OCCUR_DATE),Borough) %>%
  summarize(incidents=sum(incidents)) %>%
  filter(year==2010) %>%
  select(year, Borough, incidents)
shootings_2010
```

```{r per_capita, echo=FALSE, include=FALSE}
per_capita_2010 <- shootings_2010 %>%
  full_join(pop_2010) %>%
  mutate(shootings_per_capita = incidents/population)
per_capita_2010
```

```{r together, echo=FALSE, include=FALSE}
total_shootings_per_month <- shooting_location %>% 
  group_by(month = floor_date(OCCUR_DATE, 'month')) %>%
  summarize(incidents=sum(incidents)) %>%
  select(month, incidents)
total_shootings_per_month
summary(total_shootings_per_month$incidents)
```

```{r split_years, echo=FALSE, include=FALSE}
shootings_per_year <- shooting_location %>% 
  group_by(month = month(OCCUR_DATE),year = year(OCCUR_DATE)) %>%
  summarize(incidents=sum(incidents)) %>%
  select(month, year, incidents)
```
## Questions of Interest {data-background="newyork1.jpg}

\centering
\begin{block}{}
   \textbf{When} and \textbf{where} do the most shooting incidents in New York occur?
\end{block}

\begin{block}{}
What about the least?
\end{block}

## Cleaned and Grouped the Data

```{r show_data, echo=FALSE}
shootings_per_month
```

## Monthly Shooting Incidents in New York Boroughs

```{r incidents_over_time, echo=FALSE}
shootings_per_month %>%
  ggplot(aes(x=month, y=incidents))+
  geom_line(aes(color=Borough))+
  theme(legend.position = "bottom", 
        legend.key.size = unit(2,"line"),
        legend.text=element_text(size=13),
        legend.title=element_text(size=20),
        axis.title=element_text(size=20))+
  labs(x = "Date (By Month)",
       y = "Number of Shootings")
```

## Where do most shootings occur?

```{r pie_total, echo=FALSE}
shootings_by_borough %>%
  ggplot(aes(x="", y=incidents, fill= Borough)) +
  geom_col() +
  coord_polar(theta="y")+
  theme(legend.key.size = unit(1,"cm"),
        legend.text=element_text(size=13),
        legend.title=element_text(size=20),
        plot.title=element_text(size=25))+
  labs(title="Total Number of Shooting Incidents",
       x = "",
       y = "")
```

## Is 2010 Representative?

```{r pie_counts, echo=FALSE}
per_capita_2010 %>%
  ggplot(aes(x="", y=incidents, fill= Borough)) +
  geom_col() +
  coord_polar(theta="y")+
  theme(legend.key.size = unit(1,"cm"),
        legend.text=element_text(size=13),
        legend.title=element_text(size=20),
        plot.title=element_text(size=25))+
  labs(title="Total Number of Shooting Incidents in 2010",
       x = "",
       y = "")
```

## What if we account for population size?

```{r pie, echo=FALSE}
per_capita_2010 %>%
  ggplot(aes(x="", y=shootings_per_capita, fill= Borough)) +
  geom_col() +
  coord_polar(theta="y")+
  theme(legend.key.size = unit(1,"cm"),
        legend.text=element_text(size=13),
        legend.title=element_text(size=20),
        plot.title=element_text(size=25))+
  labs(title="Total Number of Shooting Incidents Per Capita in 2010",
       x = "",
       y = "")
```

## Monthly Spikes

```{r together_graph, echo=FALSE}
total_shootings_per_month %>%
  ggplot(aes(x=month, y=incidents))+
  geom_line(aes())+
  theme(plot.margin = margin(.1,.2,1,.2,"cm"),
        axis.title=element_text(size=20),
        plot.title=element_text(size=25))+
  labs(title= "Total Monthly Shooting Incidents in NYC",
       x = "Date (By Month)",
       y = "Number of Shootings")
```

## Total Monthly Shootings Each Year

```{r split_years_graph, echo=FALSE}
shootings_per_year %>%
  ggplot(aes(x=month, y=incidents))+
  geom_line(aes())+
  facet_wrap(~ year, ncol = 3)+
  scale_x_continuous(breaks=seq(1,12,2))+
  scale_y_continuous(breaks=seq(50,250,100))+
  theme(plot.margin = margin(.1,.2,1,.2,"cm"),
        axis.title=element_text(size=20))+
  labs(x = "Month",
       y = "Number of Shootings")
```

## NYC Shootings by Month (2006-2023)

```{r sscatter_plot, echo=FALSE}
shootings_per_year %>%
  ungroup() %>%
  ggplot(aes(x=month, y=incidents))+
  geom_point(aes())+
  scale_x_continuous(breaks=seq(1,12))+
  theme(plot.margin = margin(.1,.2,1,.2,"cm"),
        axis.title=element_text(size=20))+
  labs(x = "Month",
       y = "Number of Shootings")
```

## Model

```{r model, echo=FALSE, include=FALSE}
mod = lm(shootings_per_year$incidents ~ poly(shootings_per_year$month, degree=4,raw=T))
summary(mod)
shootings_model <- shootings_per_year %>%
  ungroup() %>%
  mutate(pred = predict(mod))
```

```{r model_graph, echo=FALSE}
shootings_model %>%
  ggplot()+
  geom_point(aes(x=month, y=incidents))+
  geom_line(aes(x=month, y=pred),color="red")+
  scale_x_continuous(breaks=seq(1,12))+
  theme(plot.margin = margin(.1,.2,1,.2,"cm"),
        axis.title=element_text(size=20))+
  labs(x = "Month",
       y = "Number of Shootings")
#shootings_model
```

## Shootings Per Month in NYC (2006-2023)

```{r boxplot, echo=FALSE}
boxplot(incidents~month, data=shootings_per_year,xlab = "Month",ylab="Number of Shootings", cex.lab = 1.45)
```

## Monthly Incident Ranges

```{r min_max_mean, echo=FALSE}
shootings_min_max_mean <- shootings_per_year %>%
  select(-c(year)) %>%
  group_by(month) %>%
  summarize(minimum = min(incidents), average = mean(incidents), maximum = max(incidents))
shootings_min_max_mean
```

## Conclusion

\begin{alertblock}{Highest per capita shooting incidents}
  \textbf{Bronx}
\end{alertblock}

\begin{exampleblock}{Lowest per capita shooting incidents}
  \textbf{Staten Island}
\end{exampleblock}

\pause

\begin{alertblock}{Month with the most shooting incidents}
  \textbf{July}
\end{alertblock}

\begin{exampleblock}{Month with the fewest shooting incidents}
  \textbf{February}
\end{exampleblock}

\pause

\begin{block}{}
  What does that mean for personal safety?
\end{block}

\begin{block}{}
  What does that mean for city planning?
\end{block}

## Source Data

NYPD shooting data

https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD